% Draft: report subsection for the resource perspective (Sections 1.6-1.8).
% PROSE DRAFT v2 (pair-author pass). Deliberately long and specific: the plan is
% to condense from here, not to expand. Every number is reproducible from
% notebooks 01/02 and the models/*.json they emit.
%
% STYLE RULE: no em dashes anywhere in this file.
%
% VIS-PLACEHOLDER comments mark figures to export. The budget realistically
% allows ONE figure in Part I; the rest are candidates for the presentation.

\subsection{Resources}

\textit{Johannes Terhorst}

The resource perspective answers three questions that the process model alone
cannot: \emph{when} is somebody there to do the work (availability), \emph{who}
is allowed to do it (permission), and \emph{which} of the allowed people
actually gets it (allocation). All three follow the same split as the rest of
the simulator. The models are mined offline in notebooks
(\texttt{01\_resource\_availability}, \texttt{02\_resource\_permissions}) with
pandas, scikit-learn and \texttt{ordinor}, and are serialised to small JSON
parameter files in \texttt{models/}. The simulation only ever \emph{loads}
JSON, which keeps it dependency-light at runtime and keeps the analysis
reproducible. Each component sits behind a narrow interface
(\texttt{PermissionModel}, an availability calendar object,
\texttt{AllocationPolicy}), so a model can be replaced without touching the
engine, the process component, or the event types.

% ─────────────────────────────────────────────────────────────────────────
\textbf{Resource availability.}

\emph{What counts as a signal that somebody was at work.} Three corrections to
the log come first, each forced by the data rather than by convention. The
timestamps are in UTC while working hours are local, so we convert to
\texttt{Europe/Amsterdam}; in UTC the workday appears to open $1.17$\,h earlier
in summer than in winter, and after conversion that seasonal swing collapses to
$0.15$\,h, which is the sanity check that the conversion was needed at all.
Only Workflow (\texttt{W\_}) events are written with a real start/complete
lifecycle: the \texttt{A\_} and \texttt{O\_} events are instantaneous
object-state milestones that only ever carry \texttt{complete}. Similar to the \ref{sec:processingtimes} the otherwise natural instinct to ``filter to \texttt{complete} events'' would
throw away all $128{,}227$ \texttt{start} events, which is the entire
availability signal. Finally, one account is not a person at all: a human
cannot be active on more than $71.4\%$ of calendar days (that is the share of
days that are weekdays), and exactly one account exceeds it, \texttt{User\_1}
at $99.2\%$. We exclude it by recomputing that structural ceiling rather than
by hardcoding its name, so the rule survives a different log.

\emph{How presence is reconstructed, and why not from spans.} A resource counts
as present on a day from its first to its last \texttt{W\_} signal \emph{on
that day}, using every transition in the lifecycle (\texttt{schedule},
\texttt{start}, \texttt{suspend}, \texttt{resume}, \texttt{complete}) rather
than only \texttt{start} and \texttt{complete}. Each transition is counted on
the date it actually occurred, and presence is never interpolated across the
gap between them. The reason is visible in the log: the intervals between
\texttt{W\_} transitions are strongly bimodal. Active work (\texttt{start} or
\texttt{resume} through to the next transition) averages $5.8$\,min and
essentially never spans a night ($0.3\%$ cross midnight), whereas waiting
(\texttt{suspend} to \texttt{resume}, in practice waiting for the customer to
respond) averages $69.9$\,h and crosses nights and weekends $55\%$ of the time.
Over the whole log that is $24{,}634$ active hours against $15.05$\,M waiting
hours, so $99.8\%$ of the elapsed \texttt{start}-to-\texttt{complete} time is
waiting rather than work. Taking a \texttt{start}-to-\texttt{complete} span as
``time at work'' would therefore drag entire weekends into the workday: a task
started on Friday and resumed on Monday would fabricate presence on Saturday
and Sunday. Counting dated transition points instead means it contributes a
Friday point and a Monday point and nothing in between. We verified this rather
than assuming it: of the $12{,}475$ work items that span two or more calendar
days, across $87{,}755$ intervening days, \emph{zero} produced presence on a
day with no real event. The same bimodality is a caveat for any
processing-time model fitted on the raw span, where the true active work is
only minutes; that is picked up in Section~1.3.

\emph{Basic model: a weekly shift profile.} For each pair of resource and
weekday we fit an opening window as $[q_{10}$ of the first event, $q_{90}$ of
the last event$]$, together with $P(\text{works that day})$, written
\texttt{p\_work}. Quantiles rather than the observed extremes: min/max reaches
$100\%$ coverage of real work events but wastes half of the modelled day
($49.4\%$ window utilisation), because a single late evening stretches the
window forever, while $q_{10}/q_{90}$ still covers $97.4\%$ of real work events
at $57.2\%$ utilisation. The comparison that justifies per-resource shifts at
all is against a single global 9--17 calendar, which covers only $80.5\%$: one
office-hours band is simply wrong for a fifth of the real work. We tile a
\emph{weekly} cycle across the assignment's two-week interval rather than
fitting an A/B week, because there is no fortnightly signal to fit: splitting
each resource's days into odd and even ISO weeks moves the median shift by
$0.36$\,h, and splitting the same days at random moves it by $0.39$\,h, so the
``fortnightly pattern'' is indistinguishable from noise.

\emph{Design decision: two tiers of resource, and where the cut sits.} A
per-weekday window is only estimable if there are enough distinct days
\emph{per weekday} behind it. We require $\geq 20$ active days
(\texttt{MIN\_ACTIVE\_DAYS}), which leaves roughly four observations per
weekday, about the minimum at which a $q_{10}/q_{90}$ pair means anything more
than a restatement of the two extreme points. That threshold is a judgement
anchored on what the estimator needs rather than the result of a sweep, and we
state it as such. It admits $121$ of the $148$ humans as \emph{Tier A}.

The remaining $27$ humans are the interesting case, and the question is what to
do with them. They are short on distinct \emph{days}, not on \emph{events}:
\texttt{User\_132}, for instance, logs $2{,}622$ events across $12$ days.
Excluding them from the model does not remove them from the simulation, it only
moves the decision to the runtime fallback, and every available fallback is
wrong in a different direction. Treating an unfitted resource as always on duty
gives a sparse resource a $24/7$ shift, which is the opposite of sparse and
lets those $27$ silently absorb all night and weekend work. Treating them as
never on duty deletes capacity the log plainly shows. We therefore fit them as
\emph{Tier B} under explicitly weaker assumptions: one \emph{pooled} window per
resource from all of its events, dropping the weekday dimension we cannot
resolve, opened only on the weekdays that resource was actually seen working.
Coarser than Tier~A, but data-driven, and it removes the need for a fallback
rather than forcing a choice between two bad ones. After this, every human with
any \texttt{W\_} signal has a window ($148$ of $148$), and only the genuine
system account (\texttt{User\_1}, which has no \texttt{W\_} lifecycle at all)
stays windowless. The model records \emph{which} accounts are automated, so the
resource component can distinguish ``automated, therefore always on duty,
because a batch process keeps no office hours'' from ``a human we failed to
fit'', instead of treating every windowless account as available around the
clock.

\emph{Advanced model: the year plan.} A weekly profile repeats forever, but a
year has days on which nobody works. \texttt{YearlyAvailability} adds two
things to it. First, a discovered public-holiday calendar: we never supply a
list of holidays, we look for weekdays on which the workforce
\emph{collectively} vanishes, since forty people are not all coincidentally ill
on the same Tuesday. Absence is judged against a \emph{local} rolling baseline
rather than a global average, because the log's headcount ramps at both ends
(staff onboarding at the start, the export truncating at the end) and a global
baseline would read those ramps as holidays. This recovers the Dutch bank
holidays unprompted (Easter Monday, King's Day, Ascension, Whit Monday, Boxing
Day) and correctly does \emph{not} flag Good Friday, which is not an official
public holiday for most Dutch employers. Second, per-resource vacation, fitted
only where it is identifiable.

\emph{The vacation subtlety.} A flat rule of ``five or more consecutive absent
working days is leave'' produces an absurd $\sim\!50$ leave days per resource
per year. The reason is that these are not full-time staff: the median modelled
resource works only $\sim\!27\%$ of weekdays, and at that rate runs of five
absent days occur roughly $15$ times a year purely by chance. The flat rule
therefore does not measure leave at all, it measures the sporadic baseline, and
then adds it to the calendar a second time on top of \texttt{p\_work}, which
already models it. Each resource consequently gets its own threshold, derived
from its own working rate: for a resource that works a weekday with probability
$q$ over $n$ working days, the expected number of absence runs of length
$\geq L$ is about $n\,q\,(1-q)^L$, and we take the smallest $L$ for which that
drops below $0.5$, so a gap counts as leave only if that resource's own rhythm
would not plausibly produce it. Beyond that, we fit leave only for
near-full-time resources ($q \geq 0.60$), and accept the consequence
deliberately: you cannot detect a fortnight off in somebody who shows up one
day in ten, so for low-rate resources leave is simply not identifiable, and
their absence stays modelled by \texttt{p\_work}, where it belongs. The data
confirms exactly where the signal dies. Grouping the blocks found \emph{without}
that cut by working rate, near-full-time staff ($>60\%$) yield $22$ blocks with
a median of $10.5$ working days and a maximum of $21$, which is what annual
leave looks like; part-time staff ($30$--$60\%$) yield $50$ blocks with a
maximum of $105$ days; and sporadic staff ($<30\%$) yield $33$ blocks with a
median of $31$ days. Five-month ``holidays'' are not holidays. The plausibility
ceiling of about four weeks therefore \emph{emerges} from the data rather than
being imposed by us: we never capped block length, we only declined to fit
leave where the signal does not exist.

\emph{What the calendar actually stores.} The deployed model keeps three
mechanisms separate rather than collapsing them into a single availability
probability, because they model different things:
\begin{itemize}
  \item \texttt{windows[r][dow]}: the hours of the day $r$ is open on that
        weekday.
  \item \texttt{p\_work[r][dow]}: the probability that $r$ works that weekday
        at all. This carries the part-time, rotating rhythm of the pool (median
        $\sim\!27\%$), and it is a \emph{per-day} draw.
  \item \texttt{vacations[r]}: a concrete set of dates on leave, sampled once
        per run. The number of blocks is drawn as
        $\text{Poisson}(n\_\text{vacations\_per\_year})$ from $r$'s own fitted
        profile, each block's length is drawn from $r$'s own observed block
        lengths, and each block is placed uniformly at random among the working
        days of the year.
\end{itemize}
Keeping \texttt{p\_work} and vacation apart is the point of the design. Leave
is \emph{contiguous}: a fortnight off is one block, not ten independent unlucky
coin flips, and folding it into a per-day probability would destroy exactly the
block structure that makes a resource unavailable for a stretch long enough to
matter to a queue. Conversely, fitting vacation for a low-rate resource would
count the same absence twice, once as a block and once through the
\texttt{p\_work} that already produced it. Holidays are placed directly, since
they are a property of the \emph{calendar}, whereas vacation is re-sampled per
resource, since it is a property of the \emph{person}: the simulation should
generate leave that is statistically like 2016's rather than replay 2016's
exact roster, which would make every run identical in its absences and would
leak the training year into the forecast. Among the $22$ resources with an
identifiable profile this yields $\sim\!24$ leave days per year, a realistic
annual allowance that was arrived at rather than assumed. The whole model
serialises to $71$\,KB of parameters. (An earlier iteration wrote $38$\,MB of
per-resource-per-day rows, and nothing in the simulation ever loaded it;
fitting parameters instead of dumping observations is what makes the model
usable by \texttt{ResourceComponent}.)

\emph{Validation (second pass).} The honest test is whether the simulated
workforce has the same rhythm as the real one. Sampling each resource in or out
of each day according to its \texttt{p\_work}, the model reproduces the weekly
rhythm, the holiday troughs, and the day-to-day variability (sd $6.8$ against
the real $7.2$), and runs $7.4\%$ light on mean headcount ($37.3$ staff per
weekday against the real $40.2$) on the stable core of the log. Folding in the
sparse resources as Tier~B is what narrows that gap from $9.2\%$ to $7.4\%$
(from $36.6$ to $37.3$); the residual is because Tier~B resources genuinely
have low \texttt{p\_work} and therefore add little, which is correct rather
than a defect. The model deliberately does not reproduce the ramps at the two
ends of the real curve, since those are onboarding and export truncation,
properties of the dataset rather than of the bank, which is why the comparison
is made on the stable core only.

% ========================================================================
% TODO (Johannes) DECISION NEEDED before the next sentence is final.
%
% The validation above samples p_work per resource per day (notebook 01,
% cell 30). The *runtime* predicate does not: YearlyAvailability.is_available()
% (analysis/availability.py:376) checks holiday, vacation and window only, and
% never consults p_work. Measured on the deployed model: at Mon 2016-06-06
% 10:00, is_available() puts 123 resources on shift, while the sum of their
% p_work(Mon) is 36.5, which is the number that matches the real 40.2. So the
% deployed calendar is ~3.3x overstaffed relative to the model we validated,
% and the conservative-error sentence below is backwards for it.
%
%   Option A (preferred): add a per-(resource, date) Bernoulli draw on p_work
%     inside is_available(), seeded deterministically from (seed, resource,
%     date) so a resource cannot flicker on and off within a single day. Then
%     the paragraph above and the sentence below both stand as written.
%   Option B: report it as a known limitation. Delete the sentence below and
%     say plainly that the deployed calendar models opening hours, holidays and
%     leave but not the part-time rhythm, so Part II runs under a workforce
%     roughly 3x larger than the real one and its contention-based results are
%     optimistic rather than conservative.
%
% This also touches Part II: occupation, k-batching and the "fire two
% employees" question all assume contention.
% ========================================================================
A $\sim\!7\%$ undercount is the honest price of those choices, and it is a
conservative error: the simulation is slightly \emph{more} resource-constrained
than reality, so queues and waiting times are, if anything, overstated rather
than optimistic.

\emph{Limitation and future work: the sparse resources belong to the full-time
teams.} Cross-referencing Tier~B against the Section~1.7 organizational model
is revealing. $22$ of the $27$ sparse resources fall into just two of the ten
groups, and they are the \emph{same} groups as the full-time
application-completion (group~3) and validation (group~6) staff, doing the same
activities. They are therefore not a separate function but an \emph{addition}
to those teams: plausibly part-timers, staff on long leave, or people on- or
offboarded mid-year. A window fitted from a handful of observations is then too
narrow in spirit, and in one case a resource was seen working for a total of
about five minutes. A more faithful model would give a sparse resource its
\emph{team's} operating window while keeping its own low \texttt{p\_work}: the
same hours as the full-timers, just not every day. We flag this as future work
rather than adopt it, because it would couple Section~1.6 to the Section~1.7
org model, and because the thinnest resources are not in the permission pool
and are therefore never allocated regardless.

% VIS-PLACEHOLDER (keep this one if only one figure fits): heat-map of the
% fitted weekly windows (open/close hour per resource x weekday) with a global
% 9-17 band overlaid, showing how individual shifts diverge from office hours.
% VIS-PLACEHOLDER (presentation): coverage vs window-utilisation sweep
% (min/max, q10/q90, q25/q75, global 9-17).
% VIS-PLACEHOLDER (presentation): simulated vs real weekday headcount over the
% log span, with the reproduced holiday troughs and the un-modelled ramps.

% ─────────────────────────────────────────────────────────────────────────
\textbf{Resource permissions (Section 1.7).}

\emph{Basic: the observed matrix.} The literal requirement is a boolean
resource\,$\times$\,activity matrix in which a pair is permitted if and only if
it was observed. It is $63\%$ dense, and it is pure memorisation. It cannot
grant a pair that is obviously sensible but never happened to occur, a third of
its entries rest on ten observations or fewer, and on a temporal hold-out it
forbids $2{,}967$ events that actually happen. That last number is the argument
for doing something better: the model is not merely coarse, it is wrong about
the future in a way we can measure.

\emph{What OrdinoR does, in plain terms.} The basic matrix asks a
one-dimensional question about a \emph{person}: has resource $r$ ever performed
activity $a$? OrdinoR~\cite{yang2022ordinor} asks about a \emph{role} instead,
and it gets there in three steps.

The first step is to describe an event by more than its activity name. Each
event is tagged with a triple called an \emph{execution context}: a case type
(something about the case, for example that it is a car loan), an activity type
(the activity itself, possibly grouped), and a time type (something about when
it happened, for example that it is a Monday). So instead of ``validate'', the
context becomes ``validate a car loan on a Monday''. Any of the three
dimensions may be a wildcard meaning ``any'', so a context can be as coarse as
the plain activity or as fine as activity and loan type and weekday together.
This is simply a knob for how finely we describe the work.

The second step is to group resources who behave alike. Resources are clustered
by which contexts they work in; the set of contexts a group's members work in
is called that group's \emph{capabilities}; and a resource inherits the
capabilities of every group it belongs to. Permission then means ``does $r$
belong to a group whose capabilities cover this context?''. That is the whole
point: it is a statement about a role rather than about a personal history,
which is what lets the model grant a sensible pair $r$ never personally did,
because $r$'s colleagues do it and therefore the group does, and lets it drop
rare one-offs, because a single stray event does not survive into the group's
profile.

The third step is how a model is scored, and there are two numbers because one
of them is gameable. \emph{Fitness} is the share of real events the model
allows, so it measures completeness. \emph{Precision} measures how tightly the
allowed behaviour matches the log, that is, how little \emph{extra} the model
permits beyond what really happens. \emph{F1} combines the two. Both are needed
because a model that simply permits everybody to do everything scores a perfect
fitness of $1.0$ at a precision near zero. Such a model is called a ``flower
model'', and it is useless: it forbids nothing, so it explains nothing. Wherever
fitness looks suspiciously perfect below, precision is what catches it.

\emph{Choosing the model: a 12-model sweep.} The three knobs map exactly onto
those three ideas, so we swept all $3\times2\times2$ combinations: how to
\emph{define} contexts (activity only, versus adding the case and time
dimensions), how to \emph{cluster} resources into groups (AHC, which yields
disjoint groups, versus MOC, which allows overlapping ones), and how to
\emph{profile} a group (FullRecall, which allows anything any member ever did,
versus OverallScore, which keeps only contexts that are frequent and shared
across the group). The sweep settles every knob empirically. FullRecall is the
flower model made concrete, at fitness $1.0$ and precision $\sim\!0.01$, and it
is discarded on precision alone. Contexts built from case type, activity type
and time type together beat activity-only, and AHC beats MOC. One finding is a
trap worth reporting: \texttt{ordinor} defaults to \emph{single} linkage, which
chains $102$ of $144$ resources into one blob and yields a meaningless
organizational model, so we use \emph{complete} linkage. The deployed
configuration is therefore \texttt{CT+AT+TT} contexts, AHC with complete
linkage, and OverallScore profiling, giving $10$ groups over $144$ resources.
As a check on our own implementation, this reproduces the OrdinoR paper's
BPIC-17 experiment at an F1 at least matching theirs on every
AHC\,+\,OverallScore row (best $0.744$ against their $0.724$), and our fast
conformance measures agree with \texttt{ordinor}'s own to $10^{-9}$.

\emph{The key design decision: deploy the causally available model, not the
best-scoring one.} The highest-F1 model in the sweep uses \emph{trace-clustered}
case types, meaning the case type is derived from a summary of how the case
behaved from beginning to end. That information cannot exist at the moment a
case needs its \emph{first} resource, because the case has not behaved yet. We
therefore deploy the best model whose case types are known at arrival, which
derives them from the loan goal, and we pay $0.025$ F1 for it. The underlying
point generalises beyond this section: a conformance score rates a model's
\emph{description} of a log that has already finished, whereas a simulation
needs a \emph{generator} that only ever looks at what is known so far. A model
can score well as a description and still be unusable as a generator.

\emph{The decisive test: does it help the simulation?} Conformance numbers rate
the permission model against the log. What we actually care about is simulated
behaviour, so we ran the simulation under each of the three models and compared
the resulting distribution of work across (resource, activity) pairs against
the real one. The org model reproduces it best, at a total-variation distance
of $0.575$, against $0.696$ for the observed matrix and $0.868$ for a hardcoded
top-20 baseline. It does so while being \emph{more} selective, permitting
roughly half as many resources per activity. It wins by being more right rather
than more permissive, which is exactly the distinction the flower model exists
to warn about.

\emph{Enforcement at runtime.} The triple costs nothing extra to evaluate. The
case type rides on the event payload, since the loan goal is sampled at arrival
anyway, and the time type comes from the clock that the Section~1.6 calendar
already requires. A model with no opinion on a dimension degrades gracefully to
a plain activity lookup, so the basic and advanced models sit behind the same
\texttt{PermissionModel} interface and are selected by a flag.

% VIS-PLACEHOLDER (presentation): fitness-precision scatter of the 12 swept
% models, highlighting the flower-model corner (FullRecall), the chosen
% CT+AT+TT / AHC / OverallScore point, and the deployed causally-available one.
% VIS-PLACEHOLDER (presentation): grouped bar of real-vs-simulated
% work-distribution TV-distance for the three permission models.

% ─────────────────────────────────────────────────────────────────────────
\textbf{Resource allocation and contention (Section 1.8 and the Part~II
baseline).}

\emph{Contention rework: a two-phase protocol.} Making resources scarce
required changing how work items start. Items are now \emph{requested} and then
\emph{started}: the process component emits \texttt{ACTIVITY\_REQUEST}, meaning
only ``this item is enabled'', and the resource component is the only thing
that emits \texttt{ACTIVITY\_START}, and only once a resource is actually
bound. This matters because the engine dispatches every event to every handler
unconditionally, so keeping a queued item in the \emph{request} state, where
the process component cannot see it, is precisely what stops it from being
executed while it is still waiting for somebody to be free. That was the fix
for the case-forking bug, in which a case could proceed down two paths at once.
It also makes waiting time fall out of the log for free, as
$\text{start} - \text{request}$.

\emph{Random allocation (the mandated baseline).} Candidates are filtered first
by the permission model (who \emph{may} do it), then by live capacity (our
\texttt{capacity} counter generalises the single-slot resource of Russell et
al.~\cite{russell2005workflow}) and by on-shift status from the Section~1.6
calendar. Among the survivors, an \texttt{AllocationPolicy} picks one. The
default \texttt{RandomPolicy} is the assignment-mandated uniform random pick
(R-RMA, pattern~15), and it is the Part~II baseline that every optimisation
policy plugs into through the same seam, without editing
\texttt{ResourceComponent} again.

\emph{Saturation (R-DE, pattern~19).} When every qualified resource is busy,
the item is deferred onto a FIFO queue and re-offered the instant a qualified
resource frees up, at $O(1)$ per release, rather than being rejected.

\emph{Adapting the heuristics to uncertain availability.} Three choices follow
from the fact that the future is not knowable at allocation time. Allocation
reads the live \texttt{busy} counter and never a forecast, because stochastic
service times make future availability genuinely unknown. The shift calendar
gates \emph{allocation} rather than \emph{execution}, so a task started at
16:55 finishes instead of being preempted at 17:00, which is what a person
would do. And an activity nobody is permitted to perform runs unassigned and is
counted in \texttt{stats()}, rather than stranding the case forever and
silently corrupting every downstream metric. Every pick uses a seeded RNG, so
the whole allocation path is bit-for-bit reproducible under a fixed seed, as
the grading requires.

% VIS-PLACEHOLDER (presentation): sequence diagram of the two-phase protocol,
% ACTIVITY_REQUEST -> permission/capacity/shift filter -> policy pick ->
% ACTIVITY_START, with the R-DE deferral loop back through RESOURCE_AVAILABLE.
